Binary companions of red giant stars with APOGEE: \\ catalog and companions that survive the common envelope

### 2018-02-20 paper/1-catalog/twoface-catalog.tex
emcee sec and refs

We perform our individual system fits---that is, make our posterior
samplings---under the following assumptions:
\begin{description}
\item[no multiplets] All radial-velocity variations of the primary star are
  induced by a single companion.
  This is motivated by the idea that triple star systems are usually
  hierarchical so that the period of the inner binary is typically much shorter
  than the orbital period of the outer body (\todo{CITATION NEEDED}).
  At present, we ignore the possibility of higher-order multiple systems.
  % \todo{revisit this - do we want to also sample over a long-term trend?}
\item[Kepler] Related to the first point, all velocity variations of the primary
  are gravitational, and we therefore ignore the possibility of coherent intrinsic variation from, e.g., stellar oscillations.
\item[SB1] All spectra are single-lined; that is, we assume that the secondary
  is significantly fainter and is thus undetected in the spectra.
  This assumption is motivated by the fact that we expect \RGB\ stars to be
  substantially more luminous than their typical companion.
  However, there are known main-sequence double-lined binary stars in the
  \apogee\ catalog \citealt{El-Badry:2018}, and an expected but unknown fraction
  of \RGB--\RGB\ binaries.
\item[simple noise model] Measurements are unbiased and noise estimates are
  correct up to an unknown excess variance.
  All noise contributions result in Gaussian uncertainties on the individual
  radial-velocity measurements.
  For upper \RGB\ stars the inferred excess variance will be a combination of
  extra systematic uncertainty and true astrophysical surface jitter
  (\todo{CITATION NEEDED}).
\end{description}

\thejoker\ is a custom-built Monte Carlo sampler designed to produce independent
posterior samples in Keplerian orbital parameters, given radial-velocity
measurements under the assumptions listed above.
Our parametrization of the orbital elements is similar to the notation in
\citet{Murray:2010}:
The radial velocity $v$ at time $t$ is given by
\begin{equation}
  v(t;\bs{\theta}) = v_0 + K\,[\cos\left(\omega + f(t; e, P, M_0)\right) +
    e\,\cos\omega]
\end{equation}
where $\bs{\theta} = (P,e,M_0,\omega,K,v_0)$---period, eccentricity, mean
anomaly at a reference time $t_0$, argument of pericenter, velocity
semi-amplitude, Barycentric velocity---and the true anomaly, $f$, is a function
of time and the specified parameters (see \sectionname~2 of
\citealt{Price-Whelan:2017} or \eqname~63 in \citealt{Murray:2010}).

@@

@@

\subsection{Initial inidivual-system posterior samplings}
\label{sec:initsampling}

For all \nstars\ \apogee\ stars with $\geq 3$ good visits (see
\sectionname~\ref{sec:data}), we use \thejoker\ to generate posterior samplings
for each star under the assumptions listed above (see
\sectionname~\ref{sec:fitting}).
We start by generating a library of \nprior\ prior samples generated under a
prior similar to that defined in \citet{Price-Whelan:2017}:
\begin{itemize}
    \item uniform or isotropic in angle parameters,
    \item uniform in log-period over the domain $[1,32768]~\textrm{day}$,
    \item a beta distribution over eccentricity with fixed parameters (\citealt{Kipping:2013}).
\end{itemize}
For the prior over the excess variance parameter, $s$, we initially use
a Gaussian over the transformed parameter $y = \ln s^2$ with the mean and
standard deviation of the initial distribution set to $(\mu_{y,0}, \sigma_{y,0})
= (10.6, 3)$ in units of $\textrm{m}~\textrm{s}^{-1}$.

We request \nposterior\ posterior samples for each source, and generate the
samples using an iterative procedure that adapts to the \todo{informativeness}
of the data to more efficiently generate posterior samples.
For sources with very constraining data, \thejoker\ may return fewer than the
requested number of samples.
When this happens, we proceed in one of two ways:
\begin{enumerate}
    \item \emph{needs MCMC}: If just 1 posterior sample is returned after
    exhausting the full library of prior samples, or if multiple (but fewer than
    \nposterior) are returned that all lie within a single mode of the posterior
    pdf, the orbit-space that is consistent with the data is treated as
    effectively unimodal. In this case, we use the location of the returned
    sample, or the median over returned samples to generate a small ball of
    initial conditions and use standard MCMC to continue sampling until we obtain \nposterior\ samples. In detail, we use an ensemble sampler
    (\citealt{Goodman:}) implemented in \python\ (\citealt{Foreman-Mackey:}),
    run with \todo{XX} walkers for \todo{XX} steps, and downsample the resulting
    MCMC chains until we have \nposterior\ samples.
    \item \emph{needs more prior samples}: If more than 1 posterior sample is
    returned after exhausting the full library of prior samples and the samples
    lie in multiple modes of the posterior pdf, the star is flagged. In this
    case, the only way to proceed is to generate more prior samples and continue
    running \thejoker\ for these sources to generate more posterior samples.
\end{description}

A full run of \thejoker\ on all \nstars\ \apogee\ stars used in this work took
approximately 300 hours on a compute cluster with 448 cores, with the time
dominated by sources with many ($\gtrsim 10$) visits.

>>>>

We perform our individual system fits---that is, make our posterior
samplings---under the following assumptions:
\begin{description}
\item[no multiplets] All radial-velocity variations of the primary star are
  induced by a single companion.
  This is motivated by the idea that triple star systems are usually
  hierarchical so that the period of the inner binary is typically much shorter
  than the orbital period of the outer body (\todo{CITATION NEEDED}).
  At present, we ignore the possibility of higher-order multiple systems.
  % \todo{revisit this - do we want to also sample over a long-term trend?}
\item[Kepler] Related to the first point, all velocity variations of the primary
  are gravitational, and we therefore ignore the possibility of coherent intrinsic variation from, e.g., stellar oscillations.
\item[SB1] All spectra are single-lined; that is, we assume that the secondary
  is significantly fainter and is thus undetected in the spectra.
  This assumption is motivated by the fact that we expect \RGB\ stars to be
  substantially more luminous than their typical companion.
  However, there are known main-sequence double-lined binary stars in the
  \apogee\ catalog \citealt{El-Badry:2018}, and an expected but unknown fraction
  of \RGB--\RGB\ binaries.
\item[simple noise model] Measurements are unbiased and noise estimates are
  correct up to an unknown excess variance.
  All noise contributions result in Gaussian uncertainties on the individual
  radial-velocity measurements.
\end{description}

\thejoker\ is a custom-built Monte Carlo sampler designed to produce independent
posterior samples in Keplerian orbital parameters, given radial-velocity
measurements under the assumptions listed above.
Our parametrization of the orbital elements is similar to the notation in
\citet{Murray:2010}:
The radial velocity $v$ at time $t$ is given by
\begin{equation}
  v(t;\bs{\theta}) = v_0 + K\,[\cos\left(\omega + f(t; e, P, M_0, t_0)\right) +
    e\,\cos\omega]
\end{equation}
where $\bs{\theta} = (P,e,M_0,\omega,K,v_0)$---period, eccentricity, mean
anomaly at a reference time $t_0$, argument of pericenter, velocity
semi-amplitude, Barycentric velocity---and the true anomaly, $f$, is a function
of time and the specified parameters (see \sectionname~2 of
\citealt{Price-Whelan:2017} or \eqname~63 in \citealt{Murray:2010}).
In addition to the orbital parameters listed above, \thejoker\ can also generate
samples in an ``excess variance'' parameter, $s^2$, that is added to the
per-visit measurement variances.
This parameter allows us to test whether the visit RV uncertainties are
underestimated
For upper \RGB\ stars the inferred excess variance will be a combination of
extra systematic uncertainty and true astrophysical surface jitter
(\todo{CITATION NEEDED}).

@@

\subsection{Initial individual-system posterior samplings}
\label{sec:initsampling}

For all \nstars\ \apogee\ stars with $\geq 3$ good visits (see
\sectionname~\ref{sec:data}), we use \thejoker\ to generate posterior samplings
for each star under the assumptions listed above (see
\sectionname~\ref{sec:fitting}).
We start by generating a library of \nprior\ prior samples generated under a
prior similar to that defined in \citet{Price-Whelan:2017}:
\begin{itemize}
    \item uniform or isotropic in angle parameters,
    \item uniform in log-period over the domain $[1,32768]~\textrm{day}$,
    \item a beta distribution over eccentricity with fixed parameters (\citealt{Kipping:2013}).
\end{itemize}
For the prior over the excess variance parameter, $s^2$, we initially use
a Gaussian over the transformed parameter $y = \ln s^2$ with the mean and
standard deviation of the initial distribution set to $(\mu_{y,0}, \sigma_{y,0})
= (10.6, 3)$ in units of $\textrm{m}~\textrm{s}^{-1}$.
The reference time for each star is set to the minimum visit epoch; $M_0$ then
becomes the mean anomaly at the first visit observation.
See \tablename~\ref{tbl:params} for a description of all parameters and priors
used.

\begin{table}[h!]
    \centering
    \begin{tabular}{ r c l }
    \hline
    name & prior & description \\
    \hline
    $P$ & $\ln P \sim \mathcal{U}(1, 32768)~\textrm{day}$ & period \\
    $e$ & $e \sim \textrm{Beta}(0.867, 3.03)$ & eccentricity \\
    $t_0$ & fixed & reference time \\
    $M_0$ & $M_0 \sim \mathcal{U}(0, 2\pi)~\textrm{rad}$ & mean anomaly at reference time \\
    $\omega$ & $\omega \sim \mathcal{U}(0, 2\pi)~\textrm{rad}$ & argument of pericenter \\
    $s^2$ & $\ln s^2 \sim \mathcal{N}(\mu_y, \sigma_y^2)$ & extra variance added to each visit variance \\
    $K$ & TODO & velocity semi-amplitude \\
    $v_0$ & TODO & system barycentric velocity \\
    \hline
    \end{tabular}
    \caption{Summary and description of parameters. $\textrm{Beta}(a, b)$ is the beta distribution with \todo{a, b}, $\mathcal{U}(a, b)$ the uniform distribution over the domain $(a, b)$, and $\mathcal{N}(\mu, \sigma^2)$ is the normal distribution with mean $\mu$ and variance $\sigma^2$.}
    \label{tbl:params}
\end{table}

We request \nposterior\ posterior samples for each source.
Depending on the data quality and phase coverage of the visits, \thejoker\ will
require different numbers of prior samples in order to rejection sample down to
the requested number of posterior samples: For few-epoch or noisy RV data, many
prior samples will pass the rejection step, whereas for very precise or
many-epoch RV data, \thejoker\ may need to go process the full library of prior
samples.
We therefore generate the posterior samples using an iterative procedure that
adaptively predicts how many prior samples to test for each star.
For sources with very constraining data, \thejoker\ may return fewer than the
requested number of samples (as few as one sample).
When this happens, we proceed in one of two ways:
\begin{enumerate}
    \item \emph{needs MCMC}: If just 1 posterior sample is returned after
    exhausting the full library of prior samples, or if multiple (but fewer than
    \nposterior) are returned that all lie within a single mode of the posterior
    pdf, the posterior pdf over orbital parameters is treated as effectively
    unimodal.
    In this case, we use the location of the returned sample, or the median over
    returned samples to generate a small ball of initial conditions and use
    standard MCMC to continue sampling until we obtain \nposterior\ samples.
    In detail, we use an ensemble sampler (\citealt{Goodman:2010}) implemented
    in \python\ (\citealt{Foreman-Mackey:2013}) for this step.
    We run with 1024 walkers for 16384 steps, take the final walker positions,
    and downsample until we have \nposterior\ samples.
    We compute convergence statistics for each run and report whether the chains
    appear converged or not.
    \item \emph{needs more prior samples}: If more than 1 posterior sample is
    returned after exhausting the full library of prior samples and the samples
    lie in multiple modes of the posterior pdf, the star is flagged. In this
    case, the only way to proceed is to generate more prior samples and continue
    running \thejoker\ for these sources to generate more posterior samples.
\end{description}

A full run of \thejoker\ on all \nstars\ \apogee\ stars used in this work took
approximately 300 hours on a compute cluster with 448 cores, with the time
dominated by sources with many ($\gtrsim 10$) visits.
\todo{XX} stars were flagged as \emph{needs MCMC}, \todo{YY} were flagged as
\emph{needs more prior samples}, and the remaining \todo{ZZ} stars completed normally.

@@
